---
title: "mapping_ranges"
format: html
date: last-modified
---

## Overview

This doc provides the code to make a map of the California coast that outlines the 100 km coastal segments, used to display our range edge analysis

## Libraries

```{r}
library(tidyverse)
library(here)
library(sf)
library(tmap)
library(lwgeom) # for st_split
```

## Data

California Coastal Commission's Coastal Zone Boundary https://map.dfg.ca.gov/metadata/ds0990.html

Table with lat long points to break up the state into 100 km chunks Currently in drive folder data_and_methods

2019 TIGER/Line® Shapefiles: Coastline https://www.census.gov/cgi-bin/geo/shapefiles/index.php?year=2019&layergroup=Coastline

Ca state shapefile https://data.ca.gov/dataset/ca-geographic-boundaries

```{r}
# Read in lat long division file 
ca_breaks <- read_csv(here('data', 'raw', 'mapping', 'CA_coast_021425.csv'))

# California coastline basemap 
ca_coastline <- st_read(here('data', 'raw', 'tl_2019_us_coastline', 'tl_2019_us_coastline.shp'))

# Filter to pacific only 
ca_coastline <- ca_coastline |> 
  filter(NAME == "Pacific")

# California state basemap 
ca_basemap <- st_read(here('data', 'raw', 'ca_state', 'CA_State.shp'))

# Dangermond shapefile 
dangermond <- st_read(here('data', 'raw', 'dangermond_shapefile', 'jldp_boundary.shp'))
```

```{r}
# CA coastal zone boundary 
ca_boundary <- st_read(here('data', 'raw', 'mapping', 'CA_coastal_zone.gdb'))

# Assign polygon labels 
ca_boundary <- ca_boundary |>
  mutate(polygon_id = row_number())

# Determine which polygons are the channel islands 
# Explore data 
tm_shape(ca_boundary) + 
    tm_polygons("polygon_id", palette = "viridis")

# Filter to mainland only 
ca_mainland <- ca_boundary |> 
  filter(polygon_id == 8)
  
```


### Explore data 
```{r}
# Explore the data 
plot(ca_mainland)

class(ca_boundary) # sf_dataframe 
st_crs(ca_boundary) #  NAD83 / California Albers 
```

Plan for map: Extend the boundaries of the coastal zones buy applying a buffer of 200 km, then clipping it to the original north and south bounds

### Buffer coastline 
```{r}
# Apply 10,000 m  buffer to coastal zones 
ca_zones_extended <- st_buffer(ca_mainland, dist = 30000)
```

```{r}
# See original boundaries and extended 
tm_shape(ca_zones_extended) + 
  tm_polygons(col = "firebrick") + 
tm_shape(ca_mainland) + 
  tm_polygons(col = "grey90")
```

```{r}
# Determine bounding box of original zones for y dimensions 
ca_zones_bbox <- st_bbox(ca_mainland)
print(ca_zones_bbox) # ymax = 450609.7 ymin = -605049.7

# Determine bounding box of extended zones for x dimensions 
extend_box <- st_bbox(ca_zones_extended)
print(extend_box) #xmax = 308316.5 xmin = -414570.2

# Create polygon of new bounds 
crop_bbox <- st_bbox(c(
  xmin = -414570.2,
  ymin = -605049.7,
  xmax = 308316.5,
  ymax = 450609.7
), crs = st_crs(ca_zones_extended))

crop_polygon <- st_as_sfc(crop_bbox)

# Crop the extended zones using the bounding box polygon
ca_zones_crop <- st_intersection(crop_polygon, ca_zones_extended)

# See new cropped extended zones 
tm_shape(ca_zones_extended) + 
  tm_polygons(col = "grey90") + # cropped 
tm_shape(ca_zones_crop) + 
  tm_polygons(col = "firebrick")  # fill size 
```


### Reproject 

Reproject to a CRS that uses lat long 

```{r}
# View CRS of data 
st_crs(ca_zones_crop)

# Transform to WGS84 which uses degrees 
ca_zones_crop <- st_transform(ca_zones_crop, crs = "WGS84")
# Check 
st_crs(ca_zones_crop)

print(st_bbox(ca_zones_crop))
```

### Split polygon at latitude specified lines 
```{r}
# Extract latitude breakpoints and sort from north to south
latitude_breaks <- sort(ca_breaks$POINT_Y, decreasing = TRUE)


# Create spliting lines from latitudes (longs from bbox of ca_zones_crop)
split_lines <- st_sfc(
  lapply(latitude_breaks, function(lat) {
    st_linestring(matrix(c(-124.8, lat, -116.7, lat), ncol = 2, byrow = TRUE))
  }),
  crs = st_crs(ca_zones_crop)
)

# Combine linestrings into single geometry 
split_lines_all <- st_union(split_lines)

# Split the coastal buffer by latitudes 
split_result <- st_split(ca_zones_crop, split_lines_all)

# Convert the resulting geometry collection into polygons
split_polygons <- st_collection_extract(split_result, "POLYGON")

# Check the class of new polygons 
class(split_polygons)

# Check the result 
tm_shape(split_polygons) + 
  tm_polygons()
```

### Label the segments with region ID
```{r}
# Locate center of each polygon 
centroids <- st_centroid(split_polygons)

# Create a data frame with region IDs based on latitude
ca_regions_df <- st_as_sf(split_polygons) |> 
  # Create region_id to indentify each region 
  mutate(region_id = rank(st_coordinates(centroids)[, 2]))  # Rank centroids by descending latitude

# View results 
tm_shape(ca_regions_df) + 
    tm_polygons("region_id", palette = "viridis") 
```

### Add base map

```{r}

st_crs(ca_coastline) == st_crs(ca_regions_df)

# Transform to match 
ca_coastline <- st_transform(ca_coastline, crs = st_crs(ca_regions_df))
ca_basemap <- st_transform(ca_basemap, crs = st_crs(ca_regions_df))
dangermond <- st_transform(dangermond, crs = st_crs(ca_regions_df))
```

### Final map

```{r}
# Bounding box to ca 100 regions 
ca_regions_bbox <- st_bbox(ca_regions_df)
# View basemap 

tm_shape(ca_basemap) + 
  tm_borders() + 
tm_shape(ca_regions_df) + 
  tm_fill(col = "grey80") + 
tm_shape(ca_regions_df) + 
  tm_borders(col = "grey40") + 
tm_shape(dangermond) + 
  tm_fill(col = "red") + 
tm_shape(ca_coastline, bbox = ca_regions_bbox) + 
  tm_lines() + 
  
  tm_scale_bar(position = c(0.02, 0.02)) + # scale bar
  tm_compass(position = c(0.01, 0.08), text.size = 0.5) # compass
```
